Conformal Counterfactual Explanations

### 2023-03-27 paper/paper.tex
more progress

Solutions to Equation~\ref{eq:general} are considered valid as soon as the predicted label matches the target label. A stripped-down counterfactual explanation is therefore little different from an adversarial example. In Figure~\ref{fig:adv}, for example, we have the baseline approach proposed in \citet{wachter2017counterfactual} to MNIST data (centre panel). This approach solves Equation~\ref{eq:general} through gradient-descent in the feature space with a penalty for the distance between the factual $x$ and the counterfactual $x^{\prime}$. The underlying classifier $M_{\theta}$ is a simple Multi-Layer Perceptron (MLP) with good test accuracy. For the generated counterfactual $x^{\prime}$ the model predicts the target label with high confidence (centre panel in Figure~\ref{fig:adv}). The explanation is valid by definition, even though it looks a lot like an Adversarial Example \citep{goodfellow2014explaining}. \citet{schut2021generating} make the connection between Adversarial Examples and Counterfactual Explanations explicit and propose using a Jacobian-Based Saliency Map Attack to solve Equation~\ref{eq:general}. They demonstrate that this approach yields realistic and sparse counterfactuals for Bayesian, adversarially robust classifiers. Applying their approach to our simple MNIST classifier does not yield a realistic counterfactual but this one, too, is valid (right panel in Figure~\ref{fig:adv}).

@@

\subsection{Predictive Uncertainty Quantification for Counterfactual Explanations}

Our proposed methodolgy is built off of the findings presented in ~\citet{schut2021generating}. The authors demonstrate that it is not only possible but remarkably easy to generate plausible counterfactuals for Black Box Models that provide predictive uncertainty estimates. By avoiding counterfactual paths that are associated with high predictive uncertainty, we end up generating counterfactuals for which the model $M_{\theta}$ predicts the target label $t$ with high confidence. Provided the model is well-calibrated, these counterfactuals are plausible which the authors demonstrate empirically through benchmarks.

\textbf{TBD}: maximizing predicted probability is equivalent to minimizing uncertainty ...

The approach proposed by \citet{schut2021generating} hinges on the crucial assumption that the Black Box Model provides predictive uncertainty estimates. The authors argue that in light of rapid advances in Bayesian Deep Learning (DL), this assumption is overall less costly than the engineering overhead induced by using surrogate models. This is even more true today, as recent work has put Laplace Approximation back on the map for truly effortless Bayesian DL \citep{immer2020improving,daxberger2021laplace,antoran2023sampling}. Nonetheless, the need for Bayesian methods may be too restrictive in some cases.

Fortunately, there is a promising alternative approach to predictive uncertainty quantification (UQ) that we will turn to next: Conformal Prediction.

@@

Conformal Prediction (CP): a scalable and statistically rigorous  Not only is CP statistically rigorous and works under minimal distributional assumptions, it is also model-agnostic and can be applied at test time. That last part is critical since it allows us to relax the assumption that the Black Box Model needs to learn to generate predictive uncertainty estimates during training. In other words, CP promises to provide a way to generate plausible counterfactuals for any model without the need for surrogate models.

@@

>>>>

Solutions to Equation~\ref{eq:general} are considered valid as soon as the predicted label matches the target label. A stripped-down counterfactual explanation is therefore little different from an adversarial example. In Figure~\ref{fig:adv}, for example, we have the baseline approach proposed in \citet{wachter2017counterfactual} to MNIST data (centre panel). This approach solves Equation~\ref{eq:general} through gradient-descent in the feature space with a penalty for the distance between the factual $x$ and the counterfactual $x^{\prime}$. The underlying classifier $M_{\theta}$ is a simple Multi-Layer Perceptron (MLP) with good test accuracy. For the generated counterfactual $x^{\prime}$ the model predicts the target label with high confidence (centre panel in Figure~\ref{fig:adv}). The explanation is valid by definition, even though it looks a lot like an Adversarial Example \citep{goodfellow2014explaining}. \citet{schut2021generating} make the connection between Adversarial Examples and Counterfactual Explanations explicit and propose using a Jacobian-Based Saliency Map Attack (JSMA) to solve Equation~\ref{eq:general}. They demonstrate that this approach yields realistic and sparse counterfactuals for Bayesian, adversarially robust classifiers. Applying their approach to our simple MNIST classifier does not yield a realistic counterfactual but this one, too, is valid (right panel in Figure~\ref{fig:adv}).

@@

\subsection{Plausible Counterfactuals through Minimal Uncertainty}

Our proposed methodology is built on the findings presented in~\citet{schut2021generating}. The authors demonstrate that it is not only possible but remarkably easy to generate plausible counterfactuals for Black Box Models that provide predictive uncertainty estimates. Their proposed algorithm solves Equation~\ref{eq:general} by greedily applying JSMA in the feature space with standard cross-entropy loss and no penalty at all. They show that this is equivalent to minimizing predictive uncertainty and hence yields counterfactuals for which the model $M_{\theta}$ predicts the target label $t$ with high confidence. Provided the model is well-calibrated, these counterfactuals are plausible which the authors demonstrate empirically through benchmarks \citep{schut2021generating}.

Unfortunately, this idea hinges on the crucial assumption that the Black Box Model provides predictive uncertainty estimates. The authors argue that in light of rapid advances in Bayesian Deep Learning (DL), this assumption is overall less costly than the engineering overhead induced by using surrogate models. This is even more true today, as recent work has put Laplace Approximation back on the map for truly effortless Bayesian DL \citep{immer2020improving,daxberger2021laplace,antoran2023sampling}. Nonetheless, the need for Bayesian methods may be too restrictive in some cases.

In looking for ways to lift that restriction, we found a promising alternative candidate for predictive uncertainty quantification (UQ) that we will briefly introduce next: Conformal Prediction.

@@

Conformal Prediction (CP) is a scalable and statistically rigorous approach to predictive UQ that works under minimal distributional assumptions \citep{angelopoulos2021gentle}. It has recently gained popularity in the Machine Learning community \citep{angelopoulos2021gentle,manokhin2022awesome}. Crucially for our intended application, CP is model-agnostic and can be applied at test time. This allows us to relax the assumption that the Black Box Model needs to learn to generate predictive uncertainty estimates during training. In other words, CP promises to provide a way to generate plausible counterfactuals for any standard discriminative model without the need for surrogate models.

...

@@

The fact that conformal classifiers produce set-valued predictions introduces a challenge: it is not immediately obvious how to use such classifiers in the context of gradient-based counterfactual search. Put differently, it is not clear how to use set-valued outputs $M_{\theta}(x)$ in Equation~\ref{eq:general}. Fortunately, \citet{stutz2022learning} have recently proposed a framework for Conformal Training that also hinges on differentiability. To evaluate the performance of conformal classifiers during training, they introduce a custom loss function as well as a smooth set size penalty.

...
